---
title: "Forecasting Daily Bike Rental Demand Using Time Series Models"
author: "Paul Trimby"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tsibble)
library(feasts)
library(fable)
library(lubridate)
library(ggtime)
```

# 1. Introduction

This report presents a time series analysis of daily bike rental demand using data from the Capital Bikeshare system (2011–2012). The objective is to explore underlying patterns in the data and develop a forecasting model capable of predicting future rental demand.

**Data Source:**  
https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset  

**Reference:**  
Fanaee-T, H., & Gama, J. (2013). *Event labeling combining ensemble detectors and background knowledge*. Progress in Artificial Intelligence.

---

# 2. Data Preparation and Exploration

## 2.1 Load the Data

```{r}
bike_data <- read_csv("day.csv")
```

## 2.2 Convert to Time Series Format

```{r}
bike_ts <- bike_data %>%
  mutate(dteday = as.Date(dteday)) %>%
  as_tsibble(index = dteday)

glimpse(bike_ts)
```

The dataset contains 731 daily observations, including total rental counts and weather-related variables.

## 2.3 Visual Exploration

```{r}
bike_ts %>%
  ggplot(aes(x = dteday, y = cnt)) +
  geom_line() +
  labs(
    title = "Daily Bike Rental Demand",
    x = "Date",
    y = "Total Rentals"
  )
```

The time plot reveals:
- An upward trend from 2011 to 2012  
- Clear weekly fluctuations  
- Increasing variability over time  

These features suggest the presence of both trend and seasonality.

---

# 3. Decomposition and Stationarity Assessment

## 3.1 STL Decomposition

```{r}
bike_decomp <- bike_ts %>%
  model(STL(cnt))

components(bike_decomp) %>%
  autoplot()
```

The STL decomposition confirms:
- A strong upward trend  
- Clear and stable weekly seasonality  
- Residual variation around these components  

## 3.2 Autocorrelation Analysis

```{r}
bike_ts %>%
  ACF(cnt) %>%
  autoplot()
```

The ACF plot shows significant autocorrelation, indicating that the series is not stationary and requires differencing before modelling.

---

# 4. ARIMA Model Fitting

## 4.1 Fit Seasonal ARIMA Model

```{r}
bike_fit <- bike_ts %>%
  model(ARIMA(cnt))

report(bike_fit)
```

The selected model:

ARIMA(1,1,1)(1,0,2)[7]

Interpretation:

- d = 1 → First differencing removes the upward trend  
- [7] → Weekly seasonality (7-day cycle)  
- D = 0 → Seasonal pattern is stable and does not require seasonal differencing  

Most coefficients are statistically significant, suggesting meaningful short-term and seasonal dependencies.

---

# 5. Model Diagnostics

```{r}
bike_fit %>%
  gg_tsresiduals()
```

Residual diagnostics indicate:

- No visible trend in residuals  
- No significant autocorrelation remaining  
- Approximately normal distribution of errors  

This suggests the model residuals behave like white noise and that the ARIMA model adequately captures the structure of the data.

---

# 6. Forecasting

```{r}
bike_fc <- bike_fit %>%
  forecast(h = "30 days")

autoplot(bike_fc, bike_ts)
```

The 30-day forecast maintains the observed weekly seasonal pattern. Prediction intervals widen over time, reflecting increasing uncertainty in longer-term forecasts.

---

# 7. Findings and Conclusion

This analysis demonstrates that daily bike rental demand exhibits both trend and strong weekly seasonality.

The ARIMA(1,1,1)(1,0,2)[7] model successfully captures:

- Long-term growth through first differencing  
- Stable weekly seasonality  
- Short-term temporal dependencies  

Residual diagnostics confirm model adequacy, and the generated forecasts provide realistic short-term projections with appropriate uncertainty estimates.

Overall, this project demonstrates a structured and diagnostic-driven time series modelling workflow, from exploration and decomposition to validation and forecasting.

---
































